---
title: "R Notebook"
output: html_notebook
---

Load the data
```{r}
nba <- read.csv("nba_final.csv")
```

Look at the data
```{r}
head(nba)
```

Select only the columns we need and clean the row with at least one NA
```{r}
nba <- nba[, c("PTS", "AST", "BLK", "TRB", "eFG.", "STL", "MP", "Pos1", "Play", "mean_views")]
nba <- na.omit(nba)
```

Look at the data again
```{r}
head(nba)
```
Summary of the data numerically
```{r}
summary(nba)
```

Summary of the response variable (mean_views)
```{r}
par(mfrow = c(1, 2))
hist(nba$mean_views, main = "Histogram of Mean Views", xlab = "Mean Views", col = "lightpink", breaks = 20)
boxplot(nba$mean_views, main = "Boxplot of Mean Views", col = "lightpink")
```

```{r}
# Histograms for continuous variables
par(mfrow = c(3, 3))

hist(nba$PTS, main = "Histogram of PTS", xlab = "Points", col = "lightblue", breaks = 20)
hist(nba$AST, main = "Histogram of AST", xlab = "Assists", col = "lightgreen", breaks = 20)
hist(nba$BLK, main = "Histogram of BLK", xlab = "Blocks", col = "lightcoral", breaks = 20)
hist(nba$TRB, main = "Histogram of TRB", xlab = "Rebounds", col = "lightyellow", breaks = 20)
hist(nba$eFG., main = "Histogram of eFG.", xlab = "Effective Field Goal %", col = "lightgray", breaks = 20)
hist(nba$STL, main = "Histogram of STL", xlab = "Steals", col = "lightcyan", breaks = 20)
hist(nba$MP, main = "Histogram of MP", xlab = "Minutes Played", col = "lavender", breaks = 20)
```

```{r}
# Boxplots for continuous variables
par(mfrow = c(1, 4))

boxplot(nba$PTS, main = "Boxplot of PTS", col = "lightblue")
boxplot(nba$AST, main = "Boxplot of AST", col = "lightgreen")
boxplot(nba$BLK, main = "Boxplot of BLK", col = "lightcoral")
boxplot(nba$TRB, main = "Boxplot of TRB", col = "lightyellow")
```

```{r}
par(mfrow = c(1, 3))
boxplot(nba$eFG., main = "Boxplot of eFG.", col = "lightgray")
boxplot(nba$STL, main = "Boxplot of STL", col = "lightcyan")
boxplot(nba$MP, main = "Boxplot of MP", col = "lavender")
```

Bar chart for categorical variables
```{r}
par(mfrow = c(1, 2))  # Arrange plots in a 1x2 grid

barplot(table(nba$Pos1), main = "Barplot of Positions", col = "lightblue", xlab = "Position", ylab = "Frequency")
barplot(table(nba$Play), main = "Barplot of Play", col = "lightgreen", xlab = "Play", ylab = "Frequency")
```

Fit a preliminary model with 5 predictors: `PTS`, `AST`, `TRB`, `MP`, AND `Play` and response variable `mean_views`
```{r}
prelim_model <- lm(mean_views ~ PTS + AST + TRB + MP + Play, nba)
summary(prelim_model)
```

Verify two conditions of using residual plots in multiple linear regression
1. Conditional mean response condition
```{r}
plot(fitted(prelim_model), nba$mean_views, xlab = "Fitted", ylab = "Actual mean_views", main = "Actual mean_views vs Fitted")
```

2. Conditional mean predictor condition
```{r}
pairs(nba[, c("PTS", "AST", "TRB", "MP")])
```

Residuals vs fitted values plot
```{r}
y_value <- resid(prelim_model)
x_value <- fitted(prelim_model)
plot(x_value, y_value, xlab = "Fitted values", ylab = "Residuals", main = "Residuals vs Fitted values")
```

Residuals vs predictor (quantitative) plot
```{r}
par(mfrow = c(2, 2))

plot(nba$PTS, y_value, xlab = "Points per game", ylab = "Residuals", main = "Residuals vs PTS (Points per game)")
plot(nba$AST, y_value, xlab = "Assists per game", ylab = "Residuals", main = "Residuals vs AST (Assists per game)")
plot(nba$TRB, y_value, xlab = "Blocks per game", ylab = "Residuals", main = "Residuals vs TRB (Total rebounds per game)")
plot(nba$MP, y_value, xlab = "Minutes played", ylab = "Residuals", main = "Residuals vs MP (Minutes played)")
```

Residuals vs each predictor (categorical) plot
```{r}
boxplot(y_value ~ nba$Play, xlab = "All-star participation", ylab = "Residuals", main = "Residuals vs Play (All-star participation)")
```

Normal QQ plot
```{r}
qqnorm(y_value)
qqline(y_value)
```
